Innovation and elaboration on the avian tree of life - supplementary materials

### 2022-08-08 Manuscript/main.tex
manuscript up (NOW MOVING TO OVERLEAF

\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1}$}\\
\noindent {\small \it 
$^1$chool of Biosciences, University of Sheffield, heffield, S10 2TN, United Kingdom.\\
$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\\}

@@

\noindent (Keywords: )\\
P
atterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity among species arises along ``an evolutionary line of least resistance'', where new phenotypes arise primarily by elaboration - evolution along this line of least resistance.
At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
The presence of distinct morphological forms suggests instead that diversity can arise via innovation - where species evolve away from the line of least resistance.
Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
Our analyses show that elaboration is a common feature at all scales, consistent with theory.
We also find that innovation is a common and major contributor to avian morphological diversity among clades.
Furthermore, we show that these patterns of innovation are replicated hierarchically throughout avian evolutionary history.
These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.

@@

Abrupt reorientation of trait space in higher taxa would be consistent with Simpsonian megaevolutionary dynamics \cite{simpson1953major}.
This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shape-space onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire groups.
We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
However, we also found various combinations of elaboration and innovation across birds.
For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
Other groups such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: add to supp?
The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig. \ref{Fig:phylogeny} and ),%{#fig_correlations}),%TODO: add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.

@@

To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2.1 CPU years and 4TB of RAM for the whole tree model and @@@ for the passeriformes model). %TODO: passeriformes model size
Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package mcmcmglmmm @@@(cite?).%TODO: cite mcmcmcglmmm

\textbf{mcmcmcglmm.}
This method consists in running relatively short monte-carlo markov chains (MCMC) across a large number of different phylogenetic trees and concatenating the resulting short posteriors into one larger one that contains the phylogenetic uncertainty information.
This was effectively done in four steps (see Fig. S\ref{Fig:mcmcmcglmm})
1) running three independent MCMCglmm chains (hereafter the parametrisation chains) for 50k generations with the traits data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\% (effectively not a very low weight on the priors);
2) extracting the following parameters from the parameterisation chains:
a) the conservative burnin average time (across the three chains) defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
b) the mini-chains priors defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
We then ran 400 mini chains in parallel with the estimated burnin time and priors in order to just run 10 samples past the burnin.
This effectively resulted in getting 10 exploitable posterior samples per tree.
We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig. S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
The whole procedure is reproducible in % TODO: add vignette @@@vignette name and implemented in the mcmcmcglmmm standalone package https://github.com/TGuillerme/mcmcmcglmmm.

\textbf{Projection/rejection analyses.}
We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that group (see Fig. S). %TODO: add figure projection
In brief we can interpret where a species or a group falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
We calculated the elaboration and innovation scores using two main approaches:
1) by clades (i.e. to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes (Fig. \ref{Fig:ellipses});
and 2) by species (i.e. to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes and b) the phylogenetic major axes of their respective clade (Fig. \ref{Fig:phylogeny}).
To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of interest and made the elaboration values absolute.
This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
The entire mathematical procedure is described in details in % TODO: link to: @@@supplementary text@@@.
The whole procedure applied to our dataset is reproducible in % TODO: link to: @@@vignette name
and implemented in the dispRity package \cite{dispRity}.

>>>>

\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
\noindent {\small \it 
$^1$chool of Biosciences, University of Sheffield, heffield, S10 2TN, United Kingdom.\\
$^2$Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\
$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}

@@

\noindent (Keywords: )\\

Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity among species arises along ``an evolutionary line of least resistance'', where new phenotypes arise primarily by elaboration - evolution along this line of least resistance.
At macro and mega-evolutionary scales, however, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general}.
The presence of distinct morphological forms suggests instead that diversity can arise via innovation - where species evolve away from the line of least resistance.
Here we apply new multi-trait methods to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
Our analyses show that elaboration is a common feature at all scales, consistent with theory.
We also find that innovation is a common and major contributor to avian morphological diversity among clades.
Furthermore, we show that these patterns of innovation are replicated hierarchically throughout avian evolutionary history.
These results suggest that both elaboration and innovation are ubiquitous from macro- to mega-evolutionary scales, and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.

@@

Abrupt reorientation of trait space in higher taxa would be consistent with Simpsonian megaevolutionary dynamics \cite{simpson1953major}.
This would lead to an expectation that species should elaborate relative to their parent clade (e.g. within orders) but show a stronger tendency towards innovation relative to the global phylogeny.
To assess the relative contributions of elaboration and innovation in beak morphology among species we estimated the algebraic projection and rejection of their position in the shape-space onto (i) the global phylogenetic major phenotypic axis, (ii) the major phenotypic axis of their respective parent superorder, and (iii) the major phenotypic axis of their respective parent order.
We did this by using the same multilinear algebraic method as described above but measuring elaboration and innovation for individual species (across the 4000 posterior samples) rather than for entire groups.
We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
However, we also found various combinations of elaboration and innovation across birds.
For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
Other groups such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig. \ref{Fig:phylogeny}). %{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.

@@

To account for phylogenetic uncertainty we needed to run the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each - the whole tree and the passeriformes ones).
Instead, we developed and implemented a ''mini chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.

\textbf{mcmcmcglmm.}
This method consists in running relatively short monte-carlo markov chains (MCMC) across a large number of different phylogenetic trees and concatenating the resulting short posteriors into one larger one that contains the phylogenetic uncertainty information.
This was effectively done in four steps (see Fig. S\ref{Fig:mcmcmcglmm})
1) running three independent MCMCglmm chains (hereafter the parametrisation chains) for 50k generations with the traits data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\% (effectively not a very low weight on the priors);
2) extracting the following parameters from the parameterisation chains:
a) the conservative burnin average time (across the three chains) defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and
b) the mini-chains priors defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
We then ran 400 mini chains in parallel with the estimated burnin time and priors in order to just run 10 samples past the burnin.
This effectively resulted in getting 10 exploitable posterior samples per tree.
We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig. S\ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain (a 400 fold improvement).
The whole procedure is reproducible in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.

\textbf{Projection/rejection analyses.}
We used the distributions of the 4000 posterior variance-covariance matrices to run the projection/rejection analyses to measure the bird beaks' elaboration and innovation scores.
In fact we can use multilinear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the variance-covariance matrices as the ''line of least resistance'', or, specifically here, the line of elaboration for each random terms (corresponding to the elaboration axes for each clade) and project either species or other major axis onto that group.
In brief we can interpret where a species or a group falls on the major axis of reference as its elaboration score and how far a species is from that major axis as its innovation score.
We calculated the elaboration and innovation scores using two main approaches:
1) by clades (i.e. to measure the elaboration/innovation between clades) where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes (Fig. \ref{Fig:ellipses});
and 2) by species (i.e. to measure the elaboration/innovation within clades) where we calculated the projections of each species onto a) the global phylogenetic major axes and b) the phylogenetic major axes of their respective clade (Fig. \ref{Fig:phylogeny}).
To make the results easier to interpret, we centered and scaled them onto the centre of the major axis of interest and made the elaboration values absolute.
This way, we can interpret an elaboration or innovation score within the 0-1 range to be not exceptional (i.e within the 95\% confidence interval of the variance-covariance matrix).
The entire mathematical procedure is described in details in the supplementary materials (section \ref{supp_projection}).
The whole procedure applied to our dataset is reproducible in \href{https://raw.rawgit.net/TGuillerme/dispRity/MCMCglmm/inst/vignettes/model_test_vignette.html}{this \texttt{dispRity} vignette} % TODO: update link to release when releasing
and implemented in the dispRity package \cite{dispRity}.
